---
title: "newest_analysis"
format: html
editor: visual
---

### Libraries

```{r}
library(tidyverse)
library(patchwork)
library(Biostrings)
```

### Paths

```{r}
### Paths

base_dir <- "//wsl.localhost/Ubuntu/home/lucio/projects/CR_pipeline"

base_dir <- normalizePath(
  base_dir,
  winslash = "/",
  mustWork = TRUE
)

path_data <- file.path(base_dir, "data")
path_intermediate <- file.path(path_data, "intermediate")
path_processed <- file.path(path_data, "proccesed")
```

### Data import

```{r}
match_data <- read_csv(
  file.path(path_processed, "iedb_match_regions_long_metadata_annotated.csv"),
  show_col_types = FALSE
)

IEDB_data <- read_csv(
  file.path(path_intermediate, "wrangled_IEDB.csv"),
  show_col_types = FALSE
)

pathogen_sequences <- read_csv(
  file.path(path_intermediate, "protein_sequences.csv"),
  show_col_types = FALSE
)

pathogen_occurrences <- read_csv(
  file.path(path_intermediate, "protein_occurrences.csv"),
  show_col_types = FALSE
)

iedb_fasta <- readAAStringSet(
  file.path(path_intermediate, "matched_iedb_source_proteins.fasta")
)
```

### Helper functions

```{r}
parse_uniprot_id <- function(x) {
  first_token <- x |>
    str_trim() |>
    str_split_i("\\s+", 1) |>
    str_remove("^.*/")

  parsed_id <- if_else(
    str_detect(first_token, fixed("|")),
    str_split_i(first_token, fixed("|"), 2),
    first_token,
    missing = NA_character_
  )

  parsed_id |>
    str_trim() |>
    str_to_upper()
}
```

### Reusable summary tables

```{r}
protein_lengths <- tibble(
  IEDB_Protein_ID = parse_uniprot_id(names(iedb_fasta)),
  Protein_Length = width(iedb_fasta)
) |>
  filter(!is.na(IEDB_Protein_ID))

conflicting_fasta_lengths <- protein_lengths |>
  group_by(IEDB_Protein_ID) |>
  summarise(
    n_lengths = n_distinct(Protein_Length),
    .groups = "drop"
  ) |>
  filter(n_lengths > 1)

if (nrow(conflicting_fasta_lengths) > 0) {
  stop(
    "Conflicting lengths were found for repeated IEDB FASTA IDs: ",
    paste(conflicting_fasta_lengths$IEDB_Protein_ID, collapse = ", ")
  )
}

protein_lengths <- protein_lengths |>
  group_by(IEDB_Protein_ID) |>
  summarise(
    Protein_Length = dplyr::first(Protein_Length),
    .groups = "drop"
  )
```

### Global filters

```{r}
histone_contaminant_ids <- c(
  "A0AB35K8W1",
  "A0AB35K7K7",
  "A0AB35K6S2"
)

high_match_iedb_ids <- c(
  "P0DMV8",
  "P10809",
  "P11021",
  "P06733",
  "P38646",
  "Q16778",
  "P62807",
  "P62805"
)

match_data_clean <- match_data |>
  mutate(
    IEDB_Protein_ID = parse_uniprot_id(IEDB_Protein_ID),
    Pathogen_Protein_ID = str_trim(str_to_upper(Pathogen_Protein_ID)),
    Proteome_ID = na_if(str_trim(as.character(Proteome_ID)), "")
  )

core_match_columns <- c(
  "IEDB_Protein_ID",
  "Pathogen_Protein_ID",
  "Matched_seq",
  "Pathogen_Start",
  "Pathogen_End",
  "IEDB_Match_Start",
  "IEDB_Match_End"
)

incomplete_core_matches <- match_data_clean |>
  filter(if_any(all_of(core_match_columns), is.na))

if (nrow(incomplete_core_matches) > 0) {
  stop(
    nrow(incomplete_core_matches),
    " match row(s) are missing fields required to define a core match."
  )
}

match_data_clean <- match_data_clean |>
  group_by(across(all_of(core_match_columns))) |>
  mutate(Match_ID = cur_group_id()) |>
  ungroup()

match_data_no_histone_contaminants <- match_data_clean |>
  filter(
    !(
      Pathogen_Protein_ID %in% histone_contaminant_ids &
        str_detect(
          coalesce(Epitope_Source, ""),
          regex("histone", ignore_case = TRUE)
        )
    )
  )

histone_filter_audit <- tibble(
  rows_removed =
    nrow(match_data_clean) - nrow(match_data_no_histone_contaminants),
  core_matches_removed =
    n_distinct(match_data_clean$Match_ID) -
      n_distinct(match_data_no_histone_contaminants$Match_ID)
)

print(histone_filter_audit)
```

### Fig 1 Distribution of matched epitopes grouped by source protein and associated disease

```{r}
# A core match supported by several assays is counted once within each assay.
assay_support <- match_data_no_histone_contaminants |>
  filter(
    !is.na(Assay_ID),
    !is.na(Epitope_Source),
    !is.na(Disease),
    !str_detect(
      Epitope_Source,
      regex("histone", ignore_case = TRUE)
    )
  ) |>
  distinct(
    Assay_ID,
    Epitope_Source,
    Disease,
    Match_ID
  )

source_disease_totals <- assay_support |>
  group_by(Epitope_Source, Disease) |>
  summarise(
    total_matches = n_distinct(Match_ID),
    .groups = "drop"
  )

assay_summary <- assay_support |>
  count(
    Assay_ID,
    Epitope_Source,
    Disease,
    name = "n_retained_matches"
  ) |>
  left_join(
    source_disease_totals,
    by = c("Epitope_Source", "Disease")
  ) |>
  mutate(
    Epitope_Disease = paste0(Epitope_Source, " (", Disease, ")"),
    Epitope_Disease = fct_reorder(
      Epitope_Disease,
      total_matches,
      .desc = TRUE
    )
  )

fig1 <- ggplot(
  assay_summary,
  aes(
    x = n_retained_matches,
    y = reorder(as.character(Assay_ID), n_retained_matches),
    fill = Epitope_Source
  )
) +
  geom_col() +
  facet_grid(
    Epitope_Disease ~ .,
    scales = "free_y",
    space = "free_y"
  ) +
  scale_fill_discrete(guide = "none") +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000),
    labels = c("0", "1", "10", "100", "1000")
  ) +
  labs(
    x = "Number of unique core matches supported by assay (log1p)",
    y = "IEDB assays"
  ) +
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0, size = 7),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.2, "lines"),
    panel.grid = element_blank(),
    axis.line.x = element_line(color = "black"),
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.5
    )
  )

fig1
```

### Fig 3 Distribution of epitope matches by pathogen protein subcellular location and rsa.

```{r}
fig3_data <- match_data_no_histone_contaminants |>
  filter(
    !is.na(Assay_ID),
    !is.na(IEDB_Protein_ID),
    !is.na(Pathogen_Organism),
    !is.na(Pathogen_Protein_ID)
  )

match_location_analysis <- fig3_data |>
  filter(!is.na(pathogen_prot_location)) |>
  group_by(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    pathogen_prot_location
  ) |>
  summarise(
    n_matches = n_distinct(Match_ID),
    n_pathogen_proteins = n_distinct(Pathogen_Protein_ID),
    n_species = n_distinct(Pathogen_Organism),
    .groups = "drop"
  )

match_rsa_analysis <- fig3_data |>
  filter(
    !is.na(rsa_pathogen_peptide_mean),
    between(rsa_pathogen_peptide_mean, 0, 1)
  ) |>
  mutate(
    rsa_group = cut(
      rsa_pathogen_peptide_mean,
      breaks = c(0, 0.25, 0.5, 0.75, Inf),
      include.lowest = TRUE,
      right = FALSE,
      labels = c(
        "0–0.25",
        "0.25–0.5",
        "0.5–0.75",
        "0.75–1"
      )
    )
  ) |>
  filter(!is.na(rsa_group)) |>
  group_by(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    rsa_group
  ) |>
  summarise(
    n_matches = n_distinct(Match_ID),
    n_pathogen_proteins = n_distinct(Pathogen_Protein_ID),
    n_species = n_distinct(Pathogen_Organism),
    .groups = "drop"
  )

fig3_theme <- theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    plot.title = element_text(
      face = "bold",
      size = 11,
      hjust = 0.5
    ),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )

p_location <- ggplot(
  match_location_analysis,
  aes(
    x = n_species,
    y = n_pathogen_proteins,
    color = pathogen_prot_location
  )
) +
  geom_point(
    position = position_jitter(
      width = 0.15,
      height = 0.15,
      seed = 123
    ),
    size = 1.5,
    alpha = 0.5
  ) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  labs(
    title = "A. Pathogen protein subcellular location",
    x = "Number of matching pathogen species",
    y = "Number of unique pathogen protein IDs",
    color = "Protein location"
  ) +
  fig3_theme

p_rsa <- ggplot(
  match_rsa_analysis,
  aes(
    x = n_species,
    y = n_pathogen_proteins,
    color = rsa_group
  )
) +
  geom_point(
    position = position_jitter(
      width = 0.15,
      height = 0.15,
      seed = 123
    ),
    size = 1.5,
    alpha = 0.5
  ) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  labs(
    title = "B. Pathogen peptide relative solvent accessibility",
    x = "Number of matching pathogen species",
    y = "Number of unique pathogen protein IDs",
    color = "RSA group"
  ) +
  fig3_theme

fig3 <- (p_location + p_rsa) + plot_layout(guides = "keep") 
fig3

remove(p_location, p_rsa)
```

### Fig 4 Heatmap showing cross-reactivity potential between epitope source proteins and pathogen species

```{r}
species_epitope_joined <- match_data_no_histone_contaminants |>
  filter(
    !IEDB_Protein_ID %in% high_match_iedb_ids,
    !is.na(Pathogen_Organism),
    !is.na(Epitope_Source)
  ) |>
  distinct(
    Pathogen_Organism,
    Epitope_Source,
    Match_ID
  ) |>
  count(
    Pathogen_Organism,
    Epitope_Source,
    name = "n_matches"
  ) |>
  mutate(
    Genus = word(Pathogen_Organism, 1),
    Species = str_replace(
      Pathogen_Organism,
      "^[^ ]+ ",
      ""
    ),
    Species = fct_rev(Species)
  )

fig4 <- ggplot(
  species_epitope_joined,
  aes(
    x = Epitope_Source,
    y = Species,
    fill = n_matches
  )
) +
  geom_tile(color = "grey80", linewidth = 0.4) +
  scale_fill_viridis_c(
    name = "Number of unique core matches",
    option = "D"
  ) +
  scale_x_discrete(
    labels = function(x) str_wrap(x, width = 30)
  ) +
  facet_grid(
    Genus ~ .,
    scales = "free_y",
    space = "free_y"
  ) +
  labs(
    x = "Epitope source",
    y = NULL
  ) +
  theme_light(base_size = 7) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5,
      size = 7
    ),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    strip.text.y = element_text(
      angle = 0,
      hjust = 0,
      face = "bold"
    ),
    panel.spacing = unit(0.3, "lines"),
    panel.grid.major.x = element_line(color = "black"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.8
    )
  )

fig4
```

### Figure S1 Distribution of total amino acid content per pathogen species in the pathogen protein dataset

```{r}
pathogen_sequences_clean <- pathogen_sequences |>
  mutate(
    Protein_ID = str_trim(str_to_upper(Protein_ID))
  )

duplicate_sequence_ids <- pathogen_sequences_clean |>
  count(Protein_ID, name = "n_rows") |>
  filter(n_rows > 1)

if (nrow(duplicate_sequence_ids) > 0) {
  stop(
    "protein_sequences.csv contains duplicated Protein_ID values: ",
    paste(duplicate_sequence_ids$Protein_ID, collapse = ", ")
  )
}

pathogen_occurrences_clean <- pathogen_occurrences |>
  mutate(
    Protein_ID = str_trim(str_to_upper(Protein_ID)),
    Proteome_ID = na_if(str_trim(as.character(Proteome_ID)), "")
  )

duplicate_occurrences <- pathogen_occurrences_clean |>
  count(Protein_ID, Proteome_ID, name = "n_rows") |>
  filter(n_rows > 1)

if (nrow(duplicate_occurrences) > 0) {
  stop(
    "protein_occurrences.csv contains duplicated Protein_ID + Proteome_ID keys."
  )
}

pathogen_data_by_species <- pathogen_occurrences_clean |>
  left_join(
    pathogen_sequences_clean |>
      select(Protein_ID, Sequence),
    by = "Protein_ID"
  )

if (any(is.na(pathogen_data_by_species$Sequence))) {
  missing_sequence_ids <- pathogen_data_by_species |>
    filter(is.na(Sequence)) |>
    distinct(Protein_ID) |>
    pull(Protein_ID)

  stop(
    "Missing sequences for pathogen occurrence IDs: ",
    paste(missing_sequence_ids, collapse = ", ")
  )
}

species_aa_summary <- pathogen_data_by_species |>
  transmute(
    Organism_Source = Genus_species,
    Proteome_ID,
    Protein_ID,
    protein_len = str_length(Sequence)
  ) |>
  filter(
    !is.na(Organism_Source),
    Organism_Source != "",
    !is.na(protein_len)
  ) |>
  distinct(
    Organism_Source,
    Proteome_ID,
    Protein_ID,
    .keep_all = TRUE
  ) |>
  group_by(Organism_Source) |>
  summarise(
    AA_len = sum(protein_len),
    n_proteomes = n_distinct(Proteome_ID),
    n_protein_occurrences = n(),
    n_unique_protein_ids = n_distinct(Protein_ID),
    .groups = "drop"
  ) |>
  arrange(desc(AA_len))


p1 <- ggplot(
  species_aa_summary,
  aes(x = AA_len)
) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black"
  ) +
  scale_x_log10(
    labels = scales::label_number(
      scale = 1e-6,
      accuracy = 0.1
    )
  ) +
  labs(
    x = "Total amino acids per species (millions, log10 scale)",
    y = "Number of species"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )


p2 <- species_aa_summary |>
  slice_max(
    order_by = AA_len,
    n = 15,
    with_ties = FALSE
  ) |>
  ggplot(
    aes(
      x = reorder(Organism_Source, AA_len),
      y = AA_len,
      fill = n_proteomes
    )
  ) +
  geom_col() +
  scale_y_continuous(
    labels = scales::label_number(scale = 1e-6)
  ) +
  scale_fill_viridis_c(option = "D") +
  coord_flip() +
  labs(
    x = "Organism",
    y = "Total amino acids (millions)",
    fill = "Number of proteomes"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )


figs1 <- p1 + p2 +
  plot_annotation(tag_levels = "A")

remove(p1, p2)

figs1
```

### Figure S2. Distribution of epitope per source protein in IEDB data.

```{r}
epitope_source_summary <- IEDB_data |>
  filter(
    !is.na(Protein_source),
    Protein_source != ""
  ) |>
  group_by(Epitope_Source = Protein_source) |>
  summarise(
    n_epitopes = n_distinct(
      Protein_ID,
      Sequence,
      epitope_start_pos,
      epitope_end_pos
    ),
    diseases = paste(
      sort(unique(Disease[!is.na(Disease) & Disease != ""])),
      collapse = ", "
    ),
    .groups = "drop"
  ) |>
  mutate(
    Epitope_Disease = if_else(
      diseases == "",
      Epitope_Source,
      paste0(Epitope_Source, "\n(", diseases, ")")
    )
  ) |>
  arrange(desc(n_epitopes))


# A: Source proteins with the most retained epitopes
p_top15 <- epitope_source_summary |>
  slice_max(
    order_by = n_epitopes,
    n = 15,
    with_ties = FALSE
  ) |>
  ggplot(aes(
    x = reorder(Epitope_Disease, n_epitopes),
    y = n_epitopes
  )) +
  geom_col(
    fill = "steelblue",
    color = "black"
  ) +
  coord_flip() +
  labs(
    x = "Epitope source protein",
    y = "Number of retained epitopes"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )


# B: Distribution across all source proteins
p_hist <- ggplot(
  epitope_source_summary,
  aes(x = n_epitopes)
) +
  geom_histogram(
    binwidth = 5,
    boundary = 0,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    x = "Number of retained epitopes per source protein",
    y = "Number of source proteins"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )


figs2 <- p_top15 + p_hist +
  plot_layout(widths = c(1.4, 1)) +
  plot_annotation(tag_levels = "A")

figs2
```

### Fig S3. Distribution of pathogen protein subcellular localization.

```{r}
figS4_data <- match_data_no_histone_contaminants |>
  distinct(
    Pathogen_Protein_ID,
    pathogen_prot_location
  ) |>
  mutate(
    pathogen_prot_location = replace_na(
      pathogen_prot_location,
      "NA"
    )
  )

figs4 <- ggplot(
  figS4_data,
  aes(x = forcats::fct_infreq(pathogen_prot_location))
) +
  geom_bar(fill = "darkseagreen3", color = "white") +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000),
    labels = c("0", "1", "10", "100", "1000")
  ) +
  labs(
    x = "Matched pathogen protein location (DeepLoc Pro)",
    y = "Number of unique pathogen protein IDs (log1p)"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

figs4
```

### Find regions of IEDB protein matched/not

```{r}
# ============================================================
# Coverage of IEDB source proteins
# ============================================================

merge_intervals <- function(intervals) {
  intervals <- intervals |>
    filter(!is.na(start), !is.na(end)) |>
    transmute(
      start = as.integer(pmin(start, end)),
      end = as.integer(pmax(start, end))
    ) |>
    distinct() |>
    arrange(start, end)

  if (nrow(intervals) == 0) {
    return(tibble(
      start = integer(),
      end = integer()
    ))
  }

  merged <- list()
  current_start <- intervals$start[1]
  current_end <- intervals$end[1]

  if (nrow(intervals) > 1) {
    for (i in 2:nrow(intervals)) {
      next_start <- intervals$start[i]
      next_end <- intervals$end[i]

      # Merge overlapping or directly adjacent intervals
      if (next_start <= current_end + 1) {
        current_end <- max(current_end, next_end)
      } else {
        merged[[length(merged) + 1]] <- tibble(
          start = current_start,
          end = current_end
        )

        current_start <- next_start
        current_end <- next_end
      }
    }
  }

  merged[[length(merged) + 1]] <- tibble(
    start = current_start,
    end = current_end
  )

  bind_rows(merged)
}


interval_length <- function(intervals) {
  if (nrow(intervals) == 0) {
    return(0L)
  }

  sum(intervals$end - intervals$start + 1)
}


intersect_intervals <- function(a, b) {
  if (nrow(a) == 0 || nrow(b) == 0) {
    return(tibble(
      start = integer(),
      end = integer()
    ))
  }

  a_named <- a |>
    transmute(
      a_start = start,
      a_end = end
    )

  b_named <- b |>
    transmute(
      b_start = start,
      b_end = end
    )

  intersections <- crossing(a_named, b_named) |>
    transmute(
      start = pmax(a_start, b_start),
      end = pmin(a_end, b_end)
    ) |>
    filter(start <= end)

  merge_intervals(intersections)
}


# ------------------------------------------------------------
# Regions covered by all IEDB epitopes included in the analysis
# ------------------------------------------------------------

studied_by_protein <- IEDB_data |>
  transmute(
    IEDB_Protein_ID = parse_uniprot_id(Protein_ID),
    Epitope_Source = Protein_source,
    Disease,
    start = epitope_start_pos,
    end = epitope_end_pos
  ) |>
  filter(
    !is.na(IEDB_Protein_ID),
    !is.na(start),
    !is.na(end)
  ) |>
  group_by(
    IEDB_Protein_ID,
    Epitope_Source,
    Disease
  ) |>
  summarise(
    studied_intervals = list(
      merge_intervals(
        tibble(
          start = start,
          end = end
        )
      )
    ),
    .groups = "drop"
  )


# ------------------------------------------------------------
# Regions with retained non-redundant exact matches
# ------------------------------------------------------------

matched_by_protein <- match_data_no_histone_contaminants |>
  mutate(
    IEDB_Protein_ID = str_to_upper(
      str_trim(IEDB_Protein_ID)
    )
  ) |>
  distinct(
    Match_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    IEDB_Match_Start,
    IEDB_Match_End
  ) |>
  group_by(
    IEDB_Protein_ID,
    Epitope_Source,
    Disease
  ) |>
  summarise(
    matched_intervals = list(
      merge_intervals(
        tibble(
          start = IEDB_Match_Start,
          end = IEDB_Match_End
        )
      )
    ),
    .groups = "drop"
  )


coverage_join_columns <- c(
  "IEDB_Protein_ID",
  "Epitope_Source",
  "Disease"
)


# Verify that every matched group can be connected to its IEDB data
missing_studied_groups <- matched_by_protein |>
  anti_join(
    studied_by_protein,
    by = coverage_join_columns
  )

if (nrow(missing_studied_groups) > 0) {
  stop(
    nrow(missing_studied_groups),
    " matched protein group(s) could not be linked to IEDB_data."
  )
}


# ------------------------------------------------------------
# Calculate the three coverage categories
# ------------------------------------------------------------

protein_coverage <- matched_by_protein |>
  inner_join(
    studied_by_protein,
    by = coverage_join_columns
  ) |>
  mutate(
    studied_length = map_int(
      studied_intervals,
      interval_length
    ),

    matched_within_studied_intervals = map2(
      studied_intervals,
      matched_intervals,
      intersect_intervals
    ),

    matched_within_studied_length = map_int(
      matched_within_studied_intervals,
      interval_length
    )
  ) |>
  left_join(
    protein_lengths,
    by = "IEDB_Protein_ID"
  )


# Stop if a source-protein length is unavailable
if (any(is.na(protein_coverage$Protein_Length))) {
  missing_length_ids <- protein_coverage |>
    filter(is.na(Protein_Length)) |>
    distinct(IEDB_Protein_ID) |>
    pull(IEDB_Protein_ID)

  stop(
    "Missing protein lengths for: ",
    paste(missing_length_ids, collapse = ", ")
  )
}


protein_coverage <- protein_coverage |>
  mutate(
    studied_but_not_matched_length =
      studied_length - matched_within_studied_length,

    not_studied_length =
      Protein_Length - studied_length,

    matched_pct =
      100 * matched_within_studied_length / Protein_Length,

    studied_but_not_matched_pct =
      100 * studied_but_not_matched_length / Protein_Length,

    not_studied_pct =
      100 * not_studied_length / Protein_Length,

    # Internal unique identifier; not displayed in the plot
    Protein_Plot_ID = paste(
      IEDB_Protein_ID,
      Epitope_Source,
      Disease,
      sep = "___"
    ),

    # Label displayed in the plot
    Protein_Label = paste0(
      Epitope_Source,
      "\n(",
      Disease,
      ")"
    )
  )


# Check for coordinates extending beyond the protein length
invalid_coverage <- protein_coverage |>
  mutate(
    studied_coordinates_invalid = map2_lgl(
      studied_intervals,
      Protein_Length,
      \(intervals, protein_length) {
        any(intervals$start < 1 | intervals$end > protein_length)
      }
    ),
    matched_coordinates_invalid = map2_lgl(
      matched_intervals,
      Protein_Length,
      \(intervals, protein_length) {
        any(intervals$start < 1 | intervals$end > protein_length)
      }
    )
  ) |>
  filter(
    studied_coordinates_invalid |
      matched_coordinates_invalid |
      studied_length > Protein_Length |
      studied_but_not_matched_length < 0 |
      not_studied_length < 0
  )

if (nrow(invalid_coverage) > 0) {
  stop(
    nrow(invalid_coverage),
    " protein group(s) have intervals inconsistent with the protein length."
  )
}


# ------------------------------------------------------------
# Convert to long format for plotting
# ------------------------------------------------------------

protein_coverage_long <- protein_coverage |>
  select(
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    Protein_Plot_ID,
    Protein_Label,
    Protein_Length,
    matched_pct,
    studied_but_not_matched_pct,
    not_studied_pct
  ) |>
  pivot_longer(
    cols = c(
      matched_pct,
      studied_but_not_matched_pct,
      not_studied_pct
    ),
    names_to = "region_type",
    values_to = "percentage"
  ) |>
  mutate(
    region_type = recode(
      region_type,
      matched_pct = "Matched",
      studied_but_not_matched_pct = "Studied but not matched",
      not_studied_pct = "Not studied"
    ),
    region_type = factor(
      region_type,
      levels = c(
        "Matched",
        "Studied but not matched",
        "Not studied"
      )
    )
  )


# Labels displayed on the y-axis
protein_plot_labels <- protein_coverage |>
  distinct(Protein_Plot_ID, Protein_Label) |>
  tibble::deframe()


# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------

ggplot(
  protein_coverage_long,
  aes(
    x = percentage,
    y = fct_reorder(Protein_Plot_ID, Protein_Length),
    fill = region_type
  )
) +
  geom_col(width = 0.8) +
  scale_x_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.01))
  ) +
  scale_y_discrete(
    labels = function(x) {
      unname(protein_plot_labels[x])
    }
  ) +
  labs(
    x = "Percentage of full IEDB source-protein length",
    y = "IEDB source protein",
    fill = "Region type"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 5)
  )
```

### Expansion of matches and analysis

```{r}
# Expansion audit
expanded_rows <- nrow(match_data_clean)
unique_core_matches <- n_distinct(match_data_clean$Match_ID)
additional_rows <- expanded_rows - unique_core_matches

expansion_audit <- tibble(
  expanded_rows = expanded_rows,
  unique_core_matches = unique_core_matches,
  expansion_factor = round(expanded_rows / unique_core_matches, 2),
  additional_rows = additional_rows,
  percentage_increase = round(
    100 * additional_rows / unique_core_matches,
    1
  )
)

print(expansion_audit)
```

### Saving neat figures

```{r}
ggsave(
  "figures/Figure_1.png",
  plot = fig1,
  width = 190,
  height = 180,
  units = "mm"
)

ggsave(
  "figures/Figure_3.png",
  plot = fig3,
  width = 180,
  height = 100,
  units = "mm"
)

ggsave(
  "figures/Figure_4.png",
  plot = fig4,
  width = 240,
  height = 150,
  units = "mm"
)

ggsave(
  "figures/Figure_S1.png",
  plot = figs1,
  width = 300,
  height = 160,
  units = "mm"
)

ggsave(
  "figures/Figure_S2.png",
  plot = figs2,
  width = 300,
  height = 160,
  units = "mm"
)

ggsave(
  "figures/Figure_S4.png",
  plot = figs4,
  width = 180,
  height = 100,
  units = "mm"
)

```

```{r}
match_rsa_analysis |>
  arrange(desc(n_matches)) |>
  dplyr::select(Epitope_Source, rsa_group, n_matches) 
```
