---
title: "epitope attributes analysis"
format: html
editor: visual
---

# Epitope attributes analysis

### Libraries:

```{r}
library("tidyverse")
library("patchwork")
library("here")
```

### Data import:

```{r}
# -----------------------------
# Main data
# -----------------------------
perfect_match <- read_csv(here("data", "proccesed", "perfect_match_DN_add.csv"),
                          show_col_types = FALSE)

pathogen_data <- read_csv(here("data", "intermediate", "protein_metadata.csv"),
                          show_col_types = FALSE)

IEDB_data <- read_csv(here("data", "intermediate", "wrangled_IEDB.csv"),
                      show_col_types = FALSE)


# -----------------------------
# Clean main data
# -----------------------------
perfect_match <- perfect_match |>
  mutate(
    Matched = as.logical(Matched),
    Epitope_Source = case_when(
      str_detect(Epitope_Source, "acetyltransferase component of pyruvate dehydrogenase complex") ~ "acetyltransferase",
      TRUE ~ Epitope_Source
    ),
    IEDB_Protein_ID = str_trim(str_to_upper(IEDB_Protein_ID)),
    Pathogen_Protein_ID = str_trim(str_to_upper(Pathogen_Protein_ID)),
    pathogen_protein_location = pathogen_prot_location,
    iedb_location = IEDB_prot_location,
    rsa_epitope = as.numeric(rsa_epitope_mean),
    rsa_pathogen_peptide = as.numeric(rsa_pathogen_peptide_mean),
    rsa_difference = rsa_pathogen_peptide - rsa_epitope
  )

pathogen_data <- pathogen_data |>
  dplyr::rename(Organism_Source = Genus_species) |>
  mutate(
    Protein_ID = str_trim(str_to_upper(Protein_ID)),
    protein_len = str_length(Sequence)
  )

IEDB_data <- IEDB_data |>
  dplyr::rename(
    IEDB_Protein_ID = Protein_ID,
    Epitope_Source = Protein_source
  ) |>
  mutate(
    IEDB_Protein_ID = str_trim(str_to_upper(IEDB_Protein_ID)),
    Epitope_Source = case_when(
      str_detect(Epitope_Source, "acetyltransferase component of pyruvate dehydrogenase complex") ~ "acetyltransferase",
      TRUE ~ Epitope_Source
    )
  )

```

### Epitope summary

```{r}
epitope_attributes <- perfect_match |>
  group_by(Assay_ID, IEDB_Protein_ID, Epitope_Source, Disease) |>
  summarise(
    epitope_location = dplyr::first(
      iedb_location[!is.na(iedb_location)],
      default = NA_character_
    ),
    matched_pathogens = list(tibble(
      Pathogen_Protein_ID = Pathogen_Protein_ID,
      pathogen_location = pathogen_protein_location,
      pathogen_species = Pathogen_Organism,
      Strain = Pathogen_Strain,
      rsa_pathogen_peptide = rsa_pathogen_peptide
    )),
    n_matches = n_distinct(
      paste(
        Pathogen_Protein_ID[!is.na(Pathogen_Protein_ID)],
        Pathogen_Strain[!is.na(Pathogen_Protein_ID)]
      )
    ),
    n_species = n_distinct(Pathogen_Organism, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(Matched = n_matches > 0)
```

### Pathogen match distribution per epitope

```{r}
# Theme: white background with black gridlines and borders
my_theme <- theme_bw(base_size = 12) +
  theme(
    panel.grid.major = element_blank(),     # remove major gridlines
    panel.grid.minor = element_blank(),     # remove minor gridlines
    panel.border = element_blank(),         # remove full border box
    axis.line = element_line(color = "black", linewidth = 0.5),  # keep x and y axis lines
    axis.text.x = element_text(angle = 40, hjust = 1)
  )

# Number of matched pathogen proteins per epitope
epitope_attributes |>
  filter(Matched) |>
  ggplot(aes(x = n_matches)) +
  geom_histogram(binwidth = 1, fill = "purple") +
  labs(
    x = "Number of matched pathogen proteins per epitope",
    y = "Number of epitopes"
  ) +
  my_theme

# Number of matched pathogen species per epitope
epitope_attributes |>
  filter(Matched) |>
  ggplot(aes(x = n_species)) +
  geom_histogram(binwidth = 1, fill = "pink", color = "white") +
  scale_y_continuous(trans = "log1p", breaks = c(0, 1, 10, 100, 1000), labels = c("0", "1", "10", "100", "1000")) +
  labs(
    x = "Number of matched pathogen species per epitope",
    y = "Number of epitopes"
  ) +
  my_theme
```

### Subcellular localization of epitope sources and matched pathogen proteins

```{r}
# Distribution of epitope subcellular locations, grouped by match status
epitope_attributes |>
  mutate(epitope_location = ifelse(is.na(epitope_location), "NA", epitope_location)) |>
  ggplot(aes(x = fct_infreq(epitope_location), fill = as.factor(Matched))) +
  geom_bar(position = "dodge", color = "white", alpha = 0.8) +
  labs(
  x = "IEDB epitope source protein location (DeepLoc)",
  y = "Number of epitopes",
  fill = "Matched to pathogen") +
  my_theme

# Distribution of pathogen protein subcellular locations (matched only)
epitope_attributes |>
  filter(Matched) |>
  unnest(matched_pathogens) |>
  distinct(Pathogen_Protein_ID, pathogen_location) |>
  mutate(pathogen_location = ifelse(is.na(pathogen_location), "NA", pathogen_location)) |>
  ggplot(aes(x = fct_infreq(pathogen_location))) +
  geom_bar(fill = "darkseagreen3", color = "white") +
  scale_y_continuous(trans = "log1p", breaks = c(0, 1, 10, 100, 1000), labels = c("0", "1", "10", "100", "1000")) +
  labs(
  x = "Matched pathogen protein location (DeepLoc Pro)",
  y = "Number of unique pathogen proteins (log10)") +
  my_theme +
  theme(
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9))
```

### Disease distribution by match status

```{r}
# Epitope count per disease, grouped by match status
epitope_attributes |>
  mutate(Disease = ifelse(is.na(Disease), "NA", Disease)) |>
  count(Disease, Matched) |>
  complete(Disease, Matched, fill = list(n = 0)) |>
  mutate(
    Disease = factor(Disease, levels = count(epitope_attributes, Disease, sort = TRUE)$Disease)
  ) |>
  ggplot(aes(x = Disease, y = n, fill = as.factor(Matched))) +
  geom_bar(stat = "identity", position = "dodge", color = "white", alpha = 0.8) +
  scale_y_continuous(trans = "log1p", breaks = c(0, 1, 10, 100),
                     labels = c("0", "1", "10", "100")) +
  labs(x = "Disease", y = "Number of Epitopes (log 10)", fill = "Matched") +
  my_theme +
  theme(
    axis.title.x = element_text(size = 9),
    axis.title.y = element_text(size = 9),
    axis.text.x = element_text(size = 6))
```

### protein location association with n_species and n_matches same with rsa

```{r}
match_location_analysis <- epitope_attributes |>
  select(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    n_matches,
    n_species,
    matched_pathogens
  ) |>
  unnest(matched_pathogens) |>
  filter(!is.na(pathogen_location)) |>
  group_by(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    pathogen_location
  ) |>
  summarise(
    n_matches = unique(n_matches),
    n_species = unique(n_species),
    count = n(),
    .groups = "drop"
  )

match_rsa_analysis <- epitope_attributes |>
  select(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    n_matches,
    n_species,
    matched_pathogens
  ) |>
  unnest(matched_pathogens) |>
  filter(!is.na(rsa_pathogen_peptide)) |>
  mutate(
    rsa_group = cut(
      rsa_pathogen_peptide,
      breaks = c(0, 0.25, 0.5, 0.75, 1),
      include.lowest = TRUE,
      right = FALSE,
      labels = c("0–0.25", "0.25–0.5", "0.5–0.75", "0.75–1")
    )
  ) |>
  group_by(
    Assay_ID,
    IEDB_Protein_ID,
    Epitope_Source,
    Disease,
    rsa_group
  ) |>
  summarise(
    n_matches = unique(n_matches),
    n_species = unique(n_species),
    count = n(),
    .groups = "drop"
  )

ggplot(
  match_location_analysis,
  aes(
    x = n_species,
    y = n_matches,
    color = pathogen_location,
    size = count
  )
) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_size_continuous(
    range = c(1, 6),
    name = "No. of Matches\nin Group",
    breaks = c(1, 10, 100, 500, 800),
    labels = c("1", "10", "100", "500", "800")
  ) +
  labs(
    x = "Number of Matching Pathogen Species (log1p scale)",
    y = "Number of Matches (log1p scale)",
    color = "Pathogen Protein Location"
  ) +
  my_theme
```

### Both plots

```{r}
library(patchwork)

p_location <- ggplot(
  match_location_analysis,
  aes(
    x = n_species,
    y = n_matches,
    color = pathogen_location,
    size = count
  )
) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_size_continuous(
    range = c(1, 6),
    name = "No. of Matches\nin Group"
  ) +
  labs(
    x = "Number of Matching Pathogen Species",
    y = "Number of Matches",
    color = "Protein Location"
  ) +
  my_theme


p_rsa <- ggplot(
  match_rsa_analysis,
  aes(
    x = n_species,
    y = n_matches,
    color = rsa_group,
    size = count
  )
) +
  geom_point(alpha = 0.5) +
  scale_x_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_y_continuous(
    trans = "log1p",
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  scale_size_continuous(
    range = c(1, 6),
    name = "No. of Matches\nin Group"
  ) +
  labs(
    x = "Number of Matching Pathogen Species",
    y = "Number of Matches",
    color = "Pathogen\nPeptide RSA"
  ) +
  my_theme

(p_location + p_rsa) +
  plot_layout(guides = "collect") &
  theme(
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
```

### IEDB epitope source composition

```{r}
# Summary: Epitope Source
epitope_summary_IEDB <- IEDB_data |>
  group_by(Epitope_Source) |>
  summarise(
    n_ep_source = n(),
    n_assay = n_distinct(Assay_ID),
    disease = dplyr::first(Disease),
    .groups = "drop"
  ) |>
  mutate(Epitope_Disease = paste0(Epitope_Source, "\n(", disease, ")")) |>
  drop_na() |>
  arrange(desc(n_ep_source))

top20_ep <- slice_max(epitope_summary_IEDB, n_ep_source, n = 20)

# Reusable plotting function
plot_bar <- function(data, x_var, y_var, fill_var = NULL,
                     x_lab = NULL, y_lab = NULL,
                     show_y_labels = TRUE, use_log = FALSE, fill_scale = NULL) {

  p <- ggplot(data, aes(x = reorder({{x_var}}, {{y_var}}), y = {{y_var}}))

  if (!rlang::quo_is_null(rlang::enquo(fill_var))) {
    p <- p + geom_bar(stat = "identity", aes(fill = {{fill_var}}))
  } else {
    p <- p + geom_bar(stat = "identity")
  }

  if (use_log) {
    p <- p + scale_y_continuous(trans = "log1p",
                                breaks = c(0, 1, 10, 100, 1000),
                                labels = c("0", "1", "10", "100", "1000"))
  }

  if (!is.null(fill_scale)) {
    p <- p + fill_scale
  }

  p +
    coord_flip() +
    labs(x = x_lab, y = y_lab) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.line = element_line(color = "black"),
      axis.text.y = if (show_y_labels) element_text() else element_blank(),
      axis.ticks.y = if (show_y_labels) element_line() else element_blank(),
      axis.text.x = element_text(angle = if (use_log) 45 else 0, hjust = 1)
    )
}

# Epitope Source Plots
p_all_ep <- plot_bar(epitope_summary_IEDB, Epitope_Source, n_ep_source,
                     x_lab = NULL, y_lab = "Number of epitopes", show_y_labels = FALSE)

p_top20_ep <- plot_bar(top20_ep, Epitope_Source, n_ep_source,
                       x_lab = "Epitope Source Protein", y_lab = "Number of epitopes")

library(patchwork)

p_top15 <- epitope_summary_IEDB |>
  slice_max(n_ep_source, n = 15) |>
  ggplot(aes(
    x = reorder(Epitope_Disease, n_ep_source),
    y = n_ep_source
  )) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    x = "Epitope Source (top 15)",
    y = "Number of Epitopes"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )

p_hist <- ggplot(epitope_summary_IEDB, aes(x = n_ep_source)) +
  geom_histogram(
    binwidth = 5,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    x = "Number of Epitopes per Source",
    y = "Number of Sources"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )

p_top15 + p_hist
```

### Match frequency across epitope source proteins

```{r}
perfect_match |>
  filter(Matched) |>
  group_by(Epitope_Source) |>
  summarise(n_matches = n(), .groups = "drop") |>
  left_join(epitope_summary_IEDB |> select(Epitope_Source, n_ep_source), by = "Epitope_Source") |>
  drop_na() |>
  arrange(desc(n_matches)) |>
  plot_bar(Epitope_Source, n_matches,
                          fill_var = n_ep_source,
                          x_lab = "Epitope Source Protein",
                          y_lab = "Frequency of Epitope source protein",
                          use_log = TRUE,
                          fill_scale = scale_fill_viridis_c(option = "D", name = "Epitopes\nin IEDB"))
```

### Match distribution across individual epitopes

```{r}
# Assay Match Frequency
assay_summary <- perfect_match |>
  filter(Matched) |>
  count(Assay_ID, Epitope_Source, Disease, name = "n_matches") |>
  group_by(Epitope_Source, Disease) |>
  mutate(total_matches = sum(n_matches)) |>
  ungroup() |>
  mutate(Epitope_Disease = paste0(Epitope_Source, " (", Disease, ")")) |>
  mutate(Epitope_Disease = fct_reorder(Epitope_Disease, total_matches, .desc = TRUE))

ggplot(assay_summary, aes(x = n_matches, y = reorder(Assay_ID, n_matches), fill = Epitope_Source)) +
  geom_col() +
  facet_grid(Epitope_Disease ~ ., scales = "free_y", space = "free_y") +
  scale_fill_discrete(guide = "none") +
  scale_x_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000),
                     labels = c("0", "1", "10", "100", "1000")) +
  labs(x = "Number of Matches (log10)", y = "Epitopes") +
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0, size = 6),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.2, "lines"),
    panel.grid = element_blank(),
    axis.line.x = element_line(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

### Pathogen species diversity across matched epitopes

```{r}
assay_species_summary <- perfect_match |>
  filter(Matched) |>
  distinct(Assay_ID, Disease, Epitope_Source, Pathogen_Organism) |>
  count(Assay_ID, Epitope_Source, Disease, name = "n_species") |>
  group_by(Epitope_Source, Disease) |>
  mutate(total_species = sum(n_species)) |>
  ungroup() |>
  mutate(Epitope_Disease = paste0(Epitope_Source, " (", Disease, ")")) |>
  mutate(Epitope_Disease = fct_reorder(Epitope_Disease, total_species, .desc = TRUE))

ggplot(assay_species_summary, aes(x = n_species, y = reorder(Assay_ID, n_species), fill = Epitope_Source)) +
  geom_col() +
  facet_grid(Epitope_Disease ~ ., scales = "free_y", space = "free_y") +
  scale_fill_discrete(guide = "none") +
  scale_x_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000),
                     labels = c("0", "1", "10", "100", "1000")) +
  labs(x = "Number of Matched Species (log)", y = "Epitopes") +
  theme_minimal() +
  theme(
    strip.text.y = element_text(angle = 0, size = 9),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.spacing = unit(0.2, "lines"),
    panel.grid = element_blank(),
    axis.line.x = element_line(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

### Disease-level distribution of IEDB epitopes and pathogen matches

```{r}
# Disease summary from IEDB
disease_summary_IEDB <- IEDB_data |>
  filter(!is.na(Disease)) |>
  count(Disease, name = "n_epitopes") |>
  arrange(desc(n_epitopes))

# Disease distribution
ggplot(disease_summary_IEDB,
       aes(x = reorder(Disease, -n_epitopes),
           y = n_epitopes)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
    x = "Autoimmune Disease",
    y = "Number of Epitopes"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    axis.text.y = element_text(),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )

# Disease Distribution Plot
disease_summary <- disease_summary_IEDB |>
  left_join(
    perfect_match |>
      filter(Matched) |>
      count(Disease, name = "n_matches"),
    by = "Disease"
  ) |>
  mutate(n_matches = replace_na(n_matches, 0))


ggplot(disease_summary, aes(x = reorder(Disease, n_matches), y = n_matches, fill = n_epitopes)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "D", name = "Epitopes in IEDB") +
  scale_y_continuous(trans = "log1p",
                     breaks = c(0, 1, 10, 100, 1000),
                     labels = c("0", "1", "10", "100", "1000")) +
  labs(
    x = "Autoimmune Disease",
    y = "Number of Matches (log10)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 8),
    axis.text.y = element_text(),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )
```

### Raw pathogen dataset composition

```{r}
# Summarize AA length and strain counts
species_aa_summary <- pathogen_data |>
  group_by(Organism_Source) |>
  summarise(
    AA_len = sum(protein_len, na.rm = TRUE),
    n_strains = n_distinct(Strain),
    .groups = "drop"
  ) |>
  arrange(desc(AA_len))


# Plot with ALL species (no organism names shown)
ggplot(species_aa_summary, aes(x = AA_len)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black") +
  scale_x_log10() +
  labs(
    x = "Total Amino Acids per Species (log10 scale)",
    y = "Number of Species with given AA content"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )


# Plot with TOP 20 species (names shown)
species_aa_summary |>
  slice_max(AA_len, n = 15) |>
  ggplot(aes(x = reorder(Organism_Source, AA_len), y = AA_len, fill = n_strains)) +
  geom_bar(stat = "identity") +
  scale_y_log10() +
  scale_fill_viridis_c(option = "D") +
  coord_flip() +
  labs(
    x = "Organism (Genus species)",
    y = "Total Amino Acids (log10 scale)",
    fill = "Number of Strains"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )
```

### Match distribution across pathogen organisms

```{r}
organism_match_stats <- perfect_match |>
  filter(Matched) |>
  group_by(Pathogen_Organism) |>
  summarise(organism_count = n(), .groups = "drop") |>
  left_join(species_aa_summary, by = c("Pathogen_Organism" = "Organism_Source")) |>
  arrange(desc(organism_count)) |>
  filter(!is.na(Pathogen_Organism)) |>
  mutate(norm_n_species = (organism_count / AA_len) * 1e5)

organism_match_stats |>
  slice_max(organism_count, n = 15) |>
  ggplot(aes(x = reorder(Pathogen_Organism, organism_count), 
             y = organism_count, fill = AA_len)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "D") +
  coord_flip() +
  labs(
    x = "Organism (Genus species)",
    y = "Number of Matches",
    fill = "Total AA's\nin Pathogen_Organism"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )

ggplot(organism_match_stats, aes(x = organism_count)) +
  geom_histogram(binwidth = 20, fill = "steelblue", color = "black") +
  labs(
    x = "Number of Matches per Organism",
    y = "Number of strain per species in pathogen data"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )
```

### Normalization of match frequency across pathogen organisms

```{r}
# Normalized match distribution across pathogens
organism_match_stats |>
  slice_max(norm_n_species, n = 20) |>
  ggplot(aes(
    x = reorder(Pathogen_Organism, norm_n_species),
    y = norm_n_species,
    fill = AA_len
  )) +
  geom_col() +
  scale_fill_viridis_c(option = "D") +
  coord_flip() +
  labs(
    x = NULL,
    y = "Matches per 100,000 amino acids",
    fill = "Total AA's\nin organism"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black")
  )
```

### Disease associations of high-matching pathogen organisms

```{r}
# Disease associations of enriched pathogens

pathogen_attributes <- perfect_match |>
  filter(Matched) |>
  group_by(Pathogen_Organism) |>
  summarise(
    epitope_sources = list(unique(Epitope_Source)),
    diseases = list(unique(Disease)),
    .groups = "drop"
  )

organism_match_stats |>
  left_join(pathogen_attributes, by = "Pathogen_Organism") |>
  arrange(desc(norm_n_species)) |>
  mutate(Pathogen_Organism = fct_inorder(Pathogen_Organism)) |>
  unnest_longer(diseases) |>
  mutate(diseases = tolower(trimws(diseases))) |>
  filter(!is.na(diseases)) |>
  filter(!diseases %in% c(
    "rheumatoid arthritis",
    "type 1 diabetes mellitus",
    "behcet's disease"
  )) |>
  mutate(present = 1) |>
  ggplot(aes(x = diseases, y = Pathogen_Organism, fill = as.factor(present))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("1" = "steelblue"), guide = "none") +
  labs(
    x = "Associated Disease",
    y = "Pathogens",
    title = "Disease Presence per Top Pathogen"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Disease-level pathogen match matrix

```{r}
# ------------------------
# 1. Disease-level matrix
# ------------------------

species_disease_raw <- perfect_match |>
  filter(Matched) |>
  filter(tolower(Disease) != "type 1 diabetes mellitus") |>
  filter(!IEDB_Protein_ID %in% c("P0DMV9", "P10809", "P11021",
                                 "P06733", "P38646")
  ) |>
  count(Pathogen_Organism, Disease, name = "n_matches")

species_disease_joined <- species_disease_raw |>
  left_join(disease_summary_IEDB, by = "Disease") |>
  left_join(organism_match_stats |> select(Pathogen_Organism, AA_len), by = "Pathogen_Organism") |>
  filter(!is.na(n_epitopes), !is.na(AA_len)) |>
  mutate(
    score_scaled = (n_matches / (AA_len)) * 1e6,
    Genus = word(Pathogen_Organism, 1),
    Species = str_replace(Pathogen_Organism, "^[^ ]+ ", ""),
    Species = fct_rev(Species)
  )

ggplot(species_disease_joined, aes(x = Disease, y = Species, fill = n_matches)) +
  geom_tile(color = "grey80", linewidth = 0.4) +
  scale_fill_viridis_c(name = "Number of Matches", option = "D") +
  facet_grid(Genus ~ ., scales = "free_y", space = "free") +
  labs(x = "Autoimmune Disease", y = NULL) +  # Remove y-axis label
  theme_light(base_size = 7) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    strip.text.y = element_text(angle = 0, hjust = 0, face = "bold"),
    panel.spacing = unit(0.3, "lines"),
    panel.grid.major.x = element_line(color = "black"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
```

### Epitope source-level pathogen match matrix

```{r}
species_epitope_raw <- perfect_match |>
  filter(Matched) |>
  filter(!IEDB_Protein_ID %in% c("P10809","P0DMV8","P11021", "P38646")) |>
  filter(!IEDB_prot_location %in% c("Mitochondrion", "Cytoplasm")) |>
  count(Pathogen_Organism, Epitope_Source, name = "n_matches")

epitope_summary_slim <- epitope_summary_IEDB |> select(Epitope_Source, n_ep_source)

species_epitope_joined <- species_epitope_raw |>
  left_join(epitope_summary_slim, by = "Epitope_Source") |>
  left_join(organism_match_stats |> select(Pathogen_Organism, AA_len), by = "Pathogen_Organism") |>
  filter(!is.na(n_ep_source), !is.na(AA_len)) |>
  mutate(
    score_scaled = (n_matches / (n_ep_source * AA_len)) * 1e6,
    Genus = word(Pathogen_Organism, 1),
    Species = str_replace(Pathogen_Organism, "^[^ ]+ ", ""),
    Species = fct_rev(Species)
  )

ggplot(species_epitope_joined, aes(x = Epitope_Source, y = Species, fill = n_matches)) +
  geom_tile(color = "grey80", linewidth = 0.4) +
  scale_fill_viridis_c(name = "Number of Matches", option = "D") +
  facet_grid(Genus ~ ., scales = "free_y", space = "free") +
  labs(x = "Epitope Source", y = NULL) +  # Remove y-axis label
  theme_light(base_size = 7) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 7),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    strip.text.y = element_text(angle = 0, hjust = 0, face = "bold"),
    panel.spacing = unit(0.3, "lines"),
    panel.grid.major.x = element_line(color = "black"),
    panel.grid.major.y = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
```

### Distribution of epitope–pathogen matches across subcellular locations

```{r}
# Raw match counts, no bias correction
match_location_raw <- perfect_match |>
  filter(Matched) |>
  filter(
    !is.na(pathogen_protein_location),
    !IEDB_Protein_ID %in% c("P0DMV9", "P10809", "P11021", "P06733", "P38646")
  ) |>
  count(Epitope_Source, pathogen_protein_location, name = "n_matches")

ggplot(match_location_raw, aes(
  x = Epitope_Source,
  y = pathogen_protein_location,
  fill = n_matches
)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(
    name = "Number of Matches",
    option = "D"
  ) +
  labs(
    x = "Epitope Source",
    y = "Subcellular Location"
  ) +
  theme_light(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)
  )


# Bias-corrected within each subcellular location
# This shows the fraction of matches within each location.
match_location_norm <- match_location_raw |>
  group_by(pathogen_protein_location) |>
  mutate(norm_matches = n_matches / sum(n_matches)) |>
  ungroup()

ggplot(match_location_norm, aes(
  x = Epitope_Source,
  y = pathogen_protein_location,
  fill = norm_matches
)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(
    name = "Fraction of Matches",
    option = "D"
  ) +
  labs(
    x = "Epitope Source",
    y = "Subcellular Location"
  ) +
  theme_light(base_size = 8) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)
  )
```

### Saving epitope attributes:

```{r}
#saveRDS(epitope_attributes, "../Data/epitope_attributes.rds")
```

### Prioritized epitope–pathogen matches with high cross-reactivity potential

```{r}


filtered_data <- perfect_match |>
  filter(
    Matched,
    !IEDB_Protein_ID %in% c("P0DMV9", "P10809", "P11021", "P06733", "P38646"),
    !is.na(pathogen_protein_location),
    pathogen_protein_location != "Cytoplasmic",
    !is.na(rsa_epitope),
    !is.na(rsa_pathogen_peptide)
  ) |>
  left_join(
    organism_match_stats |> select(Pathogen_Organism, norm_n_species),
    by = "Pathogen_Organism"
  ) |>
  arrange(desc(norm_n_species))

filtered_data |> 
  ggplot(aes(
    x = Epitope_Source,
    y = Pathogen_Organism,
    fill = pathogen_protein_location
  )) +
  geom_tile(color = "white") +
  scale_fill_brewer(palette = "Set2", name = "Subcellular Location") +
  labs(
    title = "Most Likely Cross-Reactivity Cases",
    x = "Epitope Source",
    y = "Pathogen Species"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    axis.text.y = element_text(size = 7)
  )

filtered_data |> 
  distinct(Pathogen_Organism, norm_n_species) |> 
  ggplot(aes(x = reorder(Pathogen_Organism, norm_n_species), y = norm_n_species)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top Pathogens by Normalized Match Enrichment",
    x = "Pathogen Species",
    y = "Matches per 100,000 amino acids"
  ) +
  theme_minimal()

filtered_data |>
  distinct(Disease, Pathogen_Organism, Epitope_Source) |>
  count(Disease, Epitope_Source, name = "n_pathogen_species") |>
  ggplot(aes(
    x = n_pathogen_species,
    y = reorder(Epitope_Source, n_pathogen_species)
  )) +
  geom_point(size = 3) +
  facet_wrap(~ Disease, scales = "free_y", ncol = 1) +
  labs(
    title = "Candidate pathogen matches by disease and epitope source",
    x = "Number of matched pathogen species",
    y = "Epitope source"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 7)
  )

```

### RSA plot

```{r}
library(scales)

perfect_match |>
  filter(
    Matched,
    !is.na(rsa_epitope),
    !is.na(rsa_pathogen_peptide)
  ) |>
  ggplot(aes(x = rsa_epitope, y = rsa_pathogen_peptide)) +
  geom_hex(bins = 40) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "black",
    linewidth = 0.5
  ) +
  scale_fill_viridis_c(
    option = "D",
    trans = pseudo_log_trans(sigma = 100, base = 10),
    name = "No. of matches"
  ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1), expand = FALSE) +
  labs(
    x = "Mean RSA of autoimmune epitope",
    y = "Mean RSA of matched pathogen peptide"
  ) +
  my_theme
```

### Create best candidate table

```{r}
# Step 1: Filter for matched proteins and remove unwanted proteins
filtered_data <- perfect_match |>
  filter(
    Matched,
    !IEDB_Protein_ID %in% c("P0DMV9", "P10809", "P11021", "P06733", "P38646")
  ) |>
  mutate(
    Autoimmune_Class = case_when(
      Disease %in% c("Graves' disease, Graves disease", "type 1 diabetes mellitus", "myasthenia gravis", "neuromyelitis optica", "pemphigoid gestationis", "primary biliary cholangitis", "psoriasis") ~ "Organ-specific",
      Disease %in% c("autoimmune thrombocytopenic purpura", "multiple sclerosis", "rheumatoid arthritis") ~ "Systemic",
      TRUE ~ "Unknown"
    )
  )

# Build the summary table with lists, no pathogen proteome size
summary_table <- filtered_data |>
  filter(
    pathogen_protein_location != "Cytoplasmic"
  ) |>
  group_by(Disease, Autoimmune_Class, Epitope_Source) |>
  summarise(
    Pathogen_Species = paste(unique(Pathogen_Organism), collapse = "; "),
    Pathogen_Locations = paste(unique(pathogen_protein_location), collapse = "; "),
    .groups = "drop"
  )

# Create the polished table
library(gt)
table_gt <- summary_table |>
  gt() |>
  tab_header(
    title = "Top Candidate Pathogen Matches for Autoimmune Epitopes"
  ) |>
  cols_label(
    Disease = "Autoimmune Disease",
    Autoimmune_Class = "Autoimmune Class",
    Epitope_Source = "Epitope Source",
    Pathogen_Species = "Pathogen Species",
    Pathogen_Locations = "Pathogen Protein Locations"
  ) |>
  cols_align(align = "center") |>
  opt_table_font(font = list(
    google_font("Roboto"),
    default_fonts()
  ))

# Display table
table_gt

# Save the table
#gtsave(table_gt, "../final plots/Top_Candidate_Pathogen_Matches.png")
```
